---
title: "Fiji meteorological data"
output:
  html_document:
    code_folding: hide
    fig_width: 12
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = FALSE,warning = FALSE)
```

## Bring in and format data {.tabset}

Data are from the Fiji Meteorological Service. They also directed us to SST, air temp and sea level data through the [Australian government site](http://www.bom.gov.au/oceanography/projects/spslcmp/data/monthly.shtml)

### Rainfall data

```{r cars}

library(tidyverse)
library(lubridate)
library(readxl)


## bring in all data
lakebaRF<-readxl::read_excel("data/rainfall/Lakeba_Rainfall.xlsx",skip = 1,na=c(""," "))%>%
  mutate(site="Lakeba")%>%
   pivot_longer(cols=Jan:Dec,names_to = "month",values_to = "rainfall_mm")%>%
  mutate(date=ym(paste(Year,month)))

matukuRF<-readxl::read_excel("data/rainfall/Matuku_Rainfall.xlsx",skip = 1,na=c(""," "))%>%
  mutate(site="Matuku")%>%
  pivot_longer(cols=Jan:Dec,names_to = "month",values_to = "rainfall_mm")%>%
  mutate(date=ym(paste(Year,month)))

onoILauRF<-readxl::read_excel("data/rainfall/Ono-i-Lau_Rainfall.xlsx",skip = 1,na=c(""," "))%>%
  mutate(site="Ono-i-Lau")%>%
  pivot_longer(cols=Jan:Dec,names_to = "month",values_to = "rainfall_mm")%>%
  mutate(date=ym(paste(Year,month)))

vanuabalavuRF<-readxl::read_excel("data/rainfall/Vanuabalavu_Rainfall.xlsx",skip = 1,na=c(""," "))%>%
  mutate(site="Vanuabalavu")%>%
  pivot_longer(cols=Jan:Dec,names_to = "month",values_to = "rainfall_mm")%>%
  mutate(date=ym(paste(Year,month)))


## stack data together: rainfall
fijiRF<-bind_rows(lakebaRF,matukuRF,onoILauRF,vanuabalavuRF)


## calculate ave annual rainfall
fijiRFsumm<-fijiRF%>%
  group_by(Year,site)%>%
  summarise(aveAnnRF=mean(rainfall_mm,na.rm=T))

```

#mean rainfall by site
```{r}
vanuabalavuRF
vb<-ggplot(vanuabalavuRF,aes(x=date,y=rainfall_mm))+
  geom_line(aes(color=site))+
  theme_bw()
mean("rainfall_mm")

vb

#Lakeba rainfall mean
lr<-ggplot(lakebaRF,aes(x=date,y=rainfall_mm,))+
  geom_line(aes(color=site))+
  theme_bw()
lr +
  scale_color_manual(values = c("blue"))
mean("rainfall")
lr

#Matuku rainfall mean

mr<-ggplot(matukuRF,aes(x=date,y=rainfall_mm,))+
  geom_line(aes(color=site))+
  theme_bw()
mr+
  scale_color_manual(values = c("pink"))

#OnoiLau
or<-ggplot(onoILauRF,aes(x=date,y=rainfall_mm,))+
  geom_line(aes(color=site))+
  theme_bw()
or+
  scale_color_manual(values = c("green"))



```
#My rainfall data analysis
##Stack rainfall data together
#group by site and year
```{r}
fijiRF<-bind_rows(lakebaRF,matukuRF,onoILauRF,vanuabalavuRF)

fijiRFsumm<-fijiRF%>%
  group_by(Year,site)%>%
  summarise(aveAnnRF=mean(rainfall_mm,na.rm=T))

rfPlt<-ggplot(fijiRF,aes(x=date,y=rainfall_mm,group=site))+
  geom_line(aes(color=site))+
  theme_bw()

mean("rainfall_mm")

rfPlt

```


#Plotting rainfall by month across different sites

```{r}
monthlysumm<-fijiRF%>%
  group_by(month,site)%>%
  summarise(aveAnnRF=mean(rainfall_mm,na.rm=T))

monthlysumm

ms<-ggplot(monthlysumm,aes(x=month,y=aveAnnRF,group=site))+
  geom_line(aes(color=site))+
  theme_bw()

mean(monthlysumm$aveAnnRF)

ms


```


#Jesicas analysis
#### Rainfall plots

```{r absRF_plts, echo=FALSE}

rfPlt<-ggplot(fijiRF,aes(x=date,y=rainfall_mm,group=site))+
  geom_line(aes(color=site))+
  theme_bw()

mean("rainfall_mm")

rfPlt

```
#mean annual rainfall by year across 4 sites
```{r meanAnnRF}
library(ggplot2)


aveRfPlt<-ggplot (fijiRFsumm,aes(x=Year,y=aveAnnRF,group=site))+
  geom_line(aes(color=site))+
  theme_bw()

aveRfPlt

```
#Trying to plot rainfall by  months
```{r}
aveRfPlt<-ggplot(monthlysumm,aes(x=month, y=aveAnnRF,group=site))+
  geom_line(aes(color=site))+
  theme_bw()

aveRfPlt
```
#Trying to plot for May
```{r}
 aveRfPlt<-ggplot(monthlysumm,aes(x=month, y=aveAnnRF,group=site))+
  geom_line(aes(color=site))+
  theme_bw()

aveRfPlt
```


#Cylones that fall outside of the Cyclone period 
```{r}
cyclMo<-cycl%>%
  mutate(month=month(Date),
         cyclMon=factor(month,levels=c(7,8,9,10,11,12,1,2,3,4,5,6),ordered = TRUE))
ggplot(cyclMo)+
  geom_rect(ymin=2,ymax=7,xmin=1969,xmax=2021,fill="grey80")+
  # geom_rect(ymin=11,ymax=13,xmin=1969,xmax=2021)+
  geom_point(aes(x=Year,y=cyclMon,color=damage))+
  scale_y_discrete(breaks=c(7,8,9,10,11,12,1,2,3,4,5,6))+
  scale_x_continuous(expand=c(0,0))+
  theme_bw()

```



### Cyclone data

Su pulled these data from a pdf from the Fiji Meteorological Service

Winds data (from report):

- Gale = winds between 63 - 87 km/hr (34 – 47 knots)
- Storm = winds between 88 - 117 km/hr (48 – 63 knots)
- Hurricane = winds greater than 117 km/hr (> 64 knots)

```{r}

cycl<-readxl::read_excel("data/cyclone/cycloneData.xlsx",na=c(""," "))%>%
  mutate(winds=factor(`Initial winds warning`,levels=c("Gale","Storm","Hurricane"),ordered = T))


## freq of winds by date
cyclHist<-ggplot(cycl,aes(x=Date,groups=winds,fill=winds))+
  geom_histogram(size=3)+
  theme_bw()

cyclHist

##freq of winds by year
cyclHist<-ggplot(cycl,aes(x=Year,groups=winds,fill=winds))+
  geom_histogram(size=3)+
  theme_bw()

cyclHist
```

#Extent of damage by Date
```{r}

cycl<-readxl::read_excel("data/cyclone/cycloneData.xlsx",na=c(""," "))%>%
  mutate(winds=factor(`Initial winds warning`,levels=c("Gale","Storm","Hurricane"),ordered = T),
         damage=factor(`Extent of Damage`,levels=c("Minor","Moderate","Severe"),ordered = T))
## freq
cyclHist<-ggplot(cycl,aes(x=Date,groups=damage,fill=damage))+
  geom_histogram(size=3)+
  theme_bw()
cyclHist
```


